Distance Sampling in \proglang{R}

### 2018-01-10 jss2630.Rnw
journal copy editing

\section{Introduction}
<<label=setup, echo=FALSE, include=FALSE>>=
library("knitr")
options(prompt = 'R> ')
knitr::opts_knit$set(out.format="sweave", prompt=TRUE)
knitr::opts_chunk$set(cache=TRUE)
render_sweave()
@

@@

Uncertainty in measured covariates (e.g. cluster size) and model uncertainty (when two models have similar fit but substantially different estimates) can be incorporated using the multi-analysis distance sampling package \pkg{mads} \citep{mads-pkg}. In addition, \pkg{mads} can also incorporate sightings with unknown species identification. This is done by estimating the abundance of these unidentified sightings and pro-rating them to the known species \citep{Gerrodette:2005}.

@@

\bibliography{jstatsoft}

>>>>

\section{Introduction}
<<label=setup, echo=FALSE, include=FALSE>>=
library("knitr")
options(prompt = 'R> ')
knitr::opts_knit$set(out.format="sweave", prompt=TRUE)
knitr::opts_chunk$set(cache=TRUE, fig.path="Figures/")
render_sweave()
@

@@

Uncertainty in measured covariates (e.g., cluster size) and model uncertainty (when two models have similar fit but substantially different estimates) can be incorporated using the multi-analysis distance sampling package \pkg{mads} \citep{mads-pkg}. In addition, \pkg{mads} can also incorporate sightings with unknown species identification. This is done by estimating the abundance of these unidentified sightings and pro-rating them to the known species \citep{Gerrodette:2005}.

@@

\bibliography{jss2630}

### 2018-01-10 jss2630.tex
journal copy editing

\includegraphics[width=\maxwidth]{figure/detectdists-1} \end{Schunk}
\caption{Panels show an example detection function (left), the probability density function of object distances (middle) and the resulting pdf of detection distances (right) for line transects (top row) and point transects (bottom row). The pdf of observed detection distances in the right hand plots are obtained by multiplying the detection function by the pdf of object distances and rescaling. In this example, detection probability becomes effectively zero at 500 (distances shown on $x$-axis are arbitrary).}
\label{fig:detectiondist}
\end{center}
\end{figure}

@@

\includegraphics[width=\maxwidth]{figure/points-and-lines-1} \end{Schunk}
\caption{Examples of line (top row) and point (bottom row) transect sampling. Left side plots show an example of a survey of the unit square containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals. The right side of the figure shows histograms of observed distances.}
\label{fig:pointslines}
\end{center}
\end{figure}

@@

\includegraphics[width=\maxwidth]{figure/hn-hr-par-comp-1} \end{Schunk}
\caption{Half-normal (top row) and hazard-rate (bottom row) detection functions without adjustments, varying scale ($\sigma$) and (for hazard-rate) shape ($b$) parameters (values are given above the plots). On the top row from left to right, the study species becomes more detectable (higher probability of detection at larger distances). The bottom row shows the hazard-rate model's more pronounced shoulder.}
\label{fig:hnhr}
\end{center}
\end{figure}

@@

\includegraphics[width=\maxwidth]{figure/adjust-mix-comp-1} \end{Schunk}
\caption{Possible shapes for the detection function when cosine adjustments are included for half-normal and hazard-rate models.}
\label{fig:keyadj}
\end{center}
\end{figure}

@@

\includegraphics[width=\maxwidth]{figure/minke-amakihi-hn-plot-1} \end{Schunk}
\caption{Left: fitted detection function overlayed on the histogram of observed distances for the minke whale data using half-normal model. Centre and right: plots of the probability density function for the amakihi models. Centre, hazard-rate with observer as a covariate; right, hazard-rate model with observer and minutes after sunrise as covariates. Points indicate probability of detection for a given observation (given that observation's covariate values) and lines indicate the average detection function (averaged over covariates, observer or observer+minutes after sunrise).}
\label{fig:minkeamakihi}
\end{center}
\end{figure}

@@

Detection function parameters
Scale coefficient(s):  
              estimate         se
(Intercept) 3.06441705 0.10878121
obsTJS      0.53017364 0.09956539
obsTKP      0.08885471 0.18071851

Shape coefficient(s):  
             estimate         se
(Intercept) 0.8690009 0.06261764

                        Estimate          SE         CV
Average p              0.3142723   0.0204413 0.06504326
N in covered region 3955.1685709 274.2284029 0.06933419
\end{Soutput}
\end{Schunk}
This summary information includes details of the data and model specification, as well as the values of the coefficients ($\beta_j$) and their uncertainties, an ``average'' value for the detectability (see ``Estimating abundance and variance'' for details on how this is calculated) and its uncertainty. The final line gives an estimate of abundance in the area covered by the survey (see the next section; though note this estimate does not take into account cluster size).

@@

Distance sampling Cramer-von Mises test (unweighted)
Test statistic = 0.930827 p-value = 0.00357797
\end{Soutput}
\begin{Sinput}
R> gof_ds(amakihi_hr_obs)
\end{Sinput}
\begin{Soutput}

@@

Distance sampling Cramer-von Mises test (unweighted)
Test statistic = 0.198335 p-value = 0.270719
\end{Soutput}
\end{Schunk}
We conclude that the half-normal model should be discarded. Both hazard-rate models (output only shown for hazard-rate model with observer and minutes after sunrise) had non-significant goodness of fit test statistics and are both therefore deemed plausible models. The corresponding Q-Q plots are shown in Figure \ref{amakihi-qq}, comparing the half-normal model with the hazard-rate model with observer and minutes after sunrise included.

@@

\includegraphics[width=\maxwidth]{figure/amakihiqq-1} \end{Schunk}
\caption{Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates; left) and hazard-rate model with observer as a covariate (right) for the amakihi data.}
\label{amakihi-qq}
\end{center}
\end{figure}

>>>>

\includegraphics[width=\maxwidth]{Figures/detectdists-1} \end{Schunk}
\caption{Panels show an example detection function (left), the probability density function of object distances (middle) and the resulting pdf of detection distances (right) for line transects (top row) and point transects (bottom row). The pdf of observed detection distances in the right hand plots are obtained by multiplying the detection function by the pdf of object distances and rescaling. In this example, detection probability becomes effectively zero at 500 (distances shown on $x$-axis are arbitrary).}
\label{fig:detectiondist}
\end{center}
\end{figure}

@@

\includegraphics[width=\maxwidth]{Figures/points-and-lines-1} \end{Schunk}
\caption{Examples of line (top row) and point (bottom row) transect sampling. Left side plots show an example of a survey of the unit square containing a population of 500 objects; blue dashed lines (top plot) and triangles (bottom plot) indicate sampler placement, red dots indicate detected individuals and grey dots give the locations of unobserved individuals. The right side of the figure shows histograms of observed distances.}
\label{fig:pointslines}
\end{center}
\end{figure}

@@

\includegraphics[width=\maxwidth]{Figures/hn-hr-par-comp-1} \end{Schunk}
\caption{Half-normal (top row) and hazard-rate (bottom row) detection functions without adjustments, varying scale ($\sigma$) and (for hazard-rate) shape ($b$) parameters (values are given above the plots). On the top row from left to right, the study species becomes more detectable (higher probability of detection at larger distances). The bottom row shows the hazard-rate model's more pronounced shoulder.}
\label{fig:hnhr}
\end{center}
\end{figure}

@@

\includegraphics[width=\maxwidth]{Figures/adjust-mix-comp-1} \end{Schunk}
\caption{Possible shapes for the detection function when cosine adjustments are included for half-normal and hazard-rate models.}
\label{fig:keyadj}
\end{center}
\end{figure}

@@

\includegraphics[width=\maxwidth]{Figures/minke-amakihi-hn-plot-1} \end{Schunk}
\caption{Left: fitted detection function overlayed on the histogram of observed distances for the minke whale data using half-normal model. Centre and right: plots of the probability density function for the amakihi models. Centre, hazard-rate with observer as a covariate; right, hazard-rate model with observer and minutes after sunrise as covariates. Points indicate probability of detection for a given observation (given that observation's covariate values) and lines indicate the average detection function (averaged over covariates, observer or observer+minutes after sunrise).}
\label{fig:minkeamakihi}
\end{center}
\end{figure}

@@

Detection function parameters
Scale coefficient(s):  
              estimate         se
(Intercept) 3.06441773 0.10878048
obsTJS      0.53017905 0.09956471
obsTKP      0.08885528 0.18071707

Shape coefficient(s):  
             estimate         se
(Intercept) 0.8690066 0.06261754

                       Estimate           SE         CV
Average p              0.314273   0.02044127 0.06504304
N in covered region 3955.160439 274.22700349 0.06933398
\end{Soutput}
\end{Schunk}
This summary information includes details of the data and model specification, as well as the values of the coefficients ($\beta_j$) and their uncertainties, an ``average'' value for the detectability (see ``Estimating abundance and variance'' for details on how this is calculated) and its uncertainty. The final line gives an estimate of abundance in the area covered by the survey (see the next section; though note this estimate does not take into account cluster size).

@@

Distance sampling Cramer-von Mises test (unweighted)
Test statistic = 0.930828 p-value = 0.00357795
\end{Soutput}
\begin{Sinput}
R> gof_ds(amakihi_hr_obs)
\end{Sinput}
\begin{Soutput}

@@

Distance sampling Cramer-von Mises test (unweighted)
Test statistic = 0.198333 p-value = 0.270724
\end{Soutput}
\end{Schunk}
We conclude that the half-normal model should be discarded. Both hazard-rate models (output only shown for hazard-rate model with observer and minutes after sunrise) had non-significant goodness of fit test statistics and are both therefore deemed plausible models. The corresponding Q-Q plots are shown in Figure \ref{amakihi-qq}, comparing the half-normal model with the hazard-rate model with observer and minutes after sunrise included.

@@

\includegraphics[width=\maxwidth]{Figures/amakihiqq-1} \end{Schunk}
\caption{Comparison of quantile-quantile plots for a half-normal model (no adjustments, no covariates; left) and hazard-rate model with observer as a covariate (right) for the amakihi data.}
\label{amakihi-qq}
\end{center}
\end{figure}
